\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements

### 2017-09-29 JournalPISODM2016/SubmittedV2/AnswersToReferees.tex

\documentclass[12pt,a4wide]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb} 
%\usepackage{xcolor}
\usepackage{color,soul}

%%%% OUR MACROS %%%%
\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
\newcommand{\imartin}[1]{\textcolor{magenta}{\textbf{Martin: }#1}}

\newcommand{\umberto}[1]{\marginpar{\textcolor{blue}{\textbf{Umberto: }#1}}}
\newcommand{\iumberto}[1]{\textcolor{blue}{\textbf{Umberto: }#1}}

\newcommand{\placido}[1]{\marginpar{\textcolor{purple}{\textbf{Placido: }#1}}}
\newcommand{\iplacido}[1]{\textcolor{purple}{\textbf{Placido: }#1}}
%%%%%%%%%%%

\title{$\pi$SOD-M: Building SOC Applications in the Presence of Non-Functional Requirements\\[3mm]
\textit{\large (List of changes and answers to the reviewers' observations)}}
%\author{Pl\'acido A. Souza Neto \and Genoveva Vargas-Solar \and Umberto Souza da Costa \and Martin A. Musicante}
\date{\today}

\begin{document}

\maketitle

\paragraph*{Editor's remarks:}
\begin{quotation}\sf\footnotesize

We have received the review reports for your paper "pisodm: Building SOC Applications in the Presence of Non-Functional Requirements".

We require now that you implement in your submission the following recommendations made by the reviewers:

\end{quotation}

\noindent 
All the comments have been taken into account and they are addressed in the paper. 
Our answers to the reviewers (below) explain how this has been done.

The text included in the new version is \hl{highlighted} for a better identification of the changes to the document.

\paragraph*{Reviewers' comments:}

\paragraph*{Reviewer A:} 
\begin{quotation}\sf\footnotesize
\noindent 
First of all, on the form: \\
1.	Footnotes do not appear in the paper. 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

2.	On page 21, title 5.2.4 you have a missing reference. 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

3.	On page 23, second paragraph, you have an incomplete sentence: `` \dots are implemented according a the knowledge\dots ''
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

\noindent On the content:

1.	On Section 2, the related works are cited as such. This section lucks of an evaluation of these works, and of the motivation of your proposed method. 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

2.	This method can be fully automated? Is it appropriate to non-IT professionals? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

3.	While designing the application to develop using your models, how is the selection of services done? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

4.	The paper validates the method with an industrial use case, but it lacks of a real evaluation that shows how the proposed method improves the existing once. For instance, evaluating the satisfaction of the users of the method, and comparing their satisfaction with the use of another development method. 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

5.	How the integration and the description of the constraints (Non-functional properties) improves the development process and the developed application? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

6.	These non-functional requirements can include other types of constraints? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\begin{quotation}\sf\footnotesize

7.	On page 13, you specify that the model can be automatically translated into actual computer programs. Is this actually developed? If not, how is that can be possible? 
\end{quotation}

\noindent 
\hl{TO BE COMPLETED.}

\paragraph*{Reviewer B:} 
\begin{quotation}\sf\footnotesize

Changes which must be made before publication: 
+ Footnotes do not appear 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ Some references are not inserted [?] 
\end{quotation}

\noindent 
Fixed in the new version (Section 5.2.4).

\begin{quotation}\sf\footnotesize

+ Page 4, section 3.1.1: what is "value activities"? operations on values? 
\end{quotation}

\noindent 
We added a footnote explaning the value activity concept (according to~\cite{Gordijn02valuebased}).

\begin{quotation}\sf\footnotesize

+ Page 5 , paragraph following items: e3 value value model... 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ Page 22 paragraph on the bottom of the page: a too long sentence; I do not understand anything in this paragraph. 
\end{quotation}

\noindent 
We have rewritten the paragraph.

\begin{quotation}\sf\footnotesize

+ Page 23: service composition instead of services composition 
\end{quotation}

\noindent 
Fixed in the new version.

\begin{quotation}\sf\footnotesize

+ The method is an extension of SOD-M, but I was wondering why the new name is piSOD-M. If an explanation exists, it would be interesting to indicate (since there is nothing to do with pi-calculs sometimes used in the WS world) 
\end{quotation}

\noindent 
There is no specific motivation behind the name. 
It just sounded appropriate for an extension of the existing SOD-M approach.

\begin{quotation}\sf\footnotesize

+ Section 6 does not mention any future work (except on its title!)

\end{quotation}

\noindent 
We have added a final paragraph describing directions of future work.

\section*{Final remarks}

We would like to acknowledge the anonymous referees for their useful and constructive critics to our paper.

\bibliographystyle{plain}
 \bibliography{biblio} 
\end{document}

### 2017-09-29 JournalPISODM2016/SubmittedV2/main.tex
initial version 2
initial version 2 + answers to referees

@@

\KEYWORD{MDA , Non-Functional Requirements, Service-based software process.}

@@

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional}\footnote{Functional properties of a computer system are characterized by the effects it produces when given a defined input.} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

@@

The literature emphasizes the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches assert that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnote{The study in~\cite{Ameller201542} shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}.

@@

The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}. In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.

@@

The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node.
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

@@

The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
The e$^3$value value model shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
The private application interacts with Spotify for obtaining information about the flow of music being listened by a user in return of a fee for a premium subscription. 
Finally, the private application interacts with Facebook and Twitter for updating the user's status.
We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
\end{example}

The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
In order to face this shortcoming, our method proposes the use of the BPMN meta-model as a tool for modelling this aspect of the application.

@@

BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}.
The next example illustrates the use of this meta-model.

@@

Figure~\ref{fig:CIM:usecasemetamodel} shows the \textit{$\pi$-UseCase} meta-model.
This model is  to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
The notion of \textit{policy} is used to describe NFRs.
(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnote{We use the {\sc small caps} font for referring to classes of a meta-model.}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
An {\sc End Consumer} is represented by an {\sc Actor}.
An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), as a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  by {\sc Packages}.
A {\sc Business Collaborator} denotes an external service or a system that interacts with the application that is being modeled.
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
%Thus, the services functions can be specified as being grouped into packages.

@@

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} use case.
There are three kinds of constraints:
\textit{(i)} A data pair ({\sc Value Constraint}), represented as the stereotype ``value'';
\textit{(ii)} Business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
\textit{(iii)} An {\sc Exceptional Behavior} constraint, represented by the stereotype {\sf ``exceptional\_behavior''}.
The use of these constraints is shown in the next example.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example.
We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the \textit{(i)} conditions imposed by services usage -- for example, the fact that both Facebook and Twitter require authentication in order to call their methods for updating the wall; \textit{(ii)}  conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).

@@

In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
provides meta-classes to represent workflows%\footnote{Workflows will be transformed into (implemented) service compositions.} 
 that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a single policy.
 This meta-model defines:

@@

\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
%\martin{Placido: Mudar as cores... } 
 meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules.
The conditions of each rule are evaluated.
In case of failure, the actions of the rule will be performed.
The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
 We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
Both {\em Facebook} and {\em Twitter} require authentication.
Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
%{\color{red} To be completed!.}
In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
% \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}).
The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
Both {\em Facebook} and {\em Twitter} services require authentication in order to execute methods that read and update the user data.
%A call to such services must be part of the authentication protocol required by these services.
In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
This variable is imported through the library {\sf\small OAuthPolicy.Token}.
The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: ({\sf\small R$_1$}): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and ({\sf\small R$_2$}): if the token has expired, then it is renewed.
Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
On the notification of an event of that type, a credential is obtained using the username and password.
\end{example}

@@

\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03}.% or PEWS~\cite{Placido2010LTPD}.
%
PEWS~\cite{BHM06}%,Placido2010LTPD}
 is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:

@@

\subsection{Implementation}
We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
Our tool is implemented as a series of Eclipse plug-ins: 
\begin{compactitem}
\item 	We  used the Eclipse Modeling Framework (EMF)\footnote{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models. 
From these meta-models, we  developed plug-ins to support their graphical representation.

\item	 We used  ATL\footnote{http://eclipse.org/atl/. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}
for implementing the  mappings between models.

\item 	We  used Acceleo\footnote{http://www.acceleo.org/pages/home/en} to implement  the code generation plug-in to generate executable code. 
It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
A-Policy} based service composition execution environment.
\end{compactitem}

@@

To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.%\martin{Is this a case study or a proof of concept?}
Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes, as well as to propose amendments to these practices.
The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.

@@

Figure~\ref{fig:BPMNmodel} shows the BPMN model
for the \FlyingPig\ scenario. This model is partitioned for better understanding the value exchanging process.
The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
The user is a contact member of the company, who  coordinates the assessment process.
This process  involves other members of the company as well.
%
The risk assessment process starts after a request from a company.
This corresponds to the value model, in which the start stimulus triggers the whole process.
The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnote{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}.

@@

The  PSM of our case study is given in Figure~\ref{fig:piPEWSFlyingPig}. This model is obtained by transforming the $\pi$-service composition model into a workflow.
Note that this workflow is implemented by using  BPEL constructors~\cite{BPEL} and \textit{policy} notions.

@@

Through the example we underlined that every application implements functional aspects that describe its application logic.
Recall that an application logic refers to routines that perform the activities to reach the application objective.
Also there are non functional properties derived from NFR. They refer to strategies to be considered for the application execution like: security, isolation, adaptability, atomicity, and more.
These non functional properties must be ensured at execution time, and they are not completely defined within the application logic.
%
The challenge is to define them and to associate them with the application logic considering that different to existing solutions that suppose that it is possible to access the execution stat of all the components  of an application and that the application has complete control on them, in the case for service oriented applications  the components are autonomous services
API does not necessarily export information about methods dependency (e.g., in the REST protocol);
they do not share their state (stateless).

>>>>

\usepackage{color,soul} %for color text and highlighting

@@

\KEYWORD{MDA, Non-Functional Requirements, Service-based software process.}

@@

In Service-Oriented Computing~\cite{Papazoglou2007}, pre-ex\-isting services are
combined to build an application business logic.
The selection of services is usually guided by the \textit{functional}\footnotemark \footnotetext{\hl{Functional properties of a computer system are characterized by the effects it produces when given a defined input.}} requirements of the application being developed~\cite{2,decastro1,PapazoglouH06}.
An important challenge of service-o\-rien\-ted development is  to ensure the alignment between the functional requirements imposed by the business logic and the functions actually being developed.

@@

The literature emphasizes the need for meth\-od\-ol\-o\-gies and techniques for service oriented analysis and design~\cite{Papazoglou2007}.
Existing approaches assert that the convergence of model-driven soft\-ware development, service orientation  and  busi\-ness processes improvement are key for developing quality software~\cite{watson}.
Model Driven Development (MDD)  for software systems is mainly characterized by the use of models as a product~\cite{Selic03}.
These models are successively refined from abstract specifications into actual computer programs. 
A recent literature review  concludes that current MDD approaches rarely deal with NFRs~\cite{Ameller201542}\footnotemark \footnotetext{\hl{The study shows that only 48 of the 129 papers considered in their systematic mapping deal with NFR. Moreover, most of them consider just specific aspects of NFR, while only 5 papers (out of 129) propose a more general approach.}}.

@@

The applicability of our proposal was tested with an industrial case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}. In this work we show the application of our method \pisodm to develop a service based application called \FlyingPig that provides risk assessment as a service.
The models presented here were generated as a result of interacting with software developers at GCP Global and trying to attend to all the non-functional requirements present in the application domain.

@@

The e$^3$value model  identifies the value/in\-for\-ma\-tion  exchange between system components (see Figure~\ref{fig:CIM:tpme3v}).
This model  represents a business case graphically as a set of value exchanges  ($\nabla$ $\triangle$) and value activities (rounded boxes)\footnotemark \footnotetext{\hl{According to}~\cite{Gordijn02valuebased}\hl{, a value activity is defined as an activity expected to be profitable to one or more actors.}} performed by business actors (squared boxes).
The model is well suited to enhance the understanding of the environment in which an application is being developed.
It defines \textit{dependency paths}, showing the value exchange between providers and end users when they ask for a service.
A dependency path has a direction and consists of a sequence of linked dependency nodes.
It starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node.
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

@@

The e$^3$value model for ``To Publish Music'' is shown in Figure~\ref{fig:CIM:tpme3v}.
The e$^3$\hl{value model} shows Spotify and a private application (which is also a service) that provide services for listening and publishing information about music being listened by users. 
The private application interacts with Spotify for obtaining information about the flow of music being listened by a user in return of a fee for a premium subscription. 
Finally, the private application interacts with Facebook and Twitter for updating the user's status.
We can see this interaction as non material benefit sharing (as users subscribe to their networks and are active on them thanks to the private application).
\end{example}

The e$^3$value  model is used to model the (economic) value exchange among the actors involved in an application.
However, the e$^3$val\-ue model does not help to understand the business process of the application nor the conditions in which the different steps of this process are executed.
In order to face this shortcoming, our method proposes the use of the BPMN meta-model as a tool for modeling this aspect of the application.

@@

BPMN~\cite{BPMN}  is a graphical representation that establishes the business process of the application through a high-level workflow\footnotemark \footnotetext{\hl{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/}}.
The next example illustrates the use of this meta-model.

@@

Figure~\ref{fig:CIM:usecasemetamodel} shows the \textit{$\pi$-UseCase} meta-model.
This model is  to be the first representation of the application in terms of functionality, as well as to represent its non-func\-tion\-al requirements.
The notion of \textit{policy} is used to describe NFRs.
(Figure~\ref{fig:CIM:usecasemetamodel} highlights the concepts related to NFRs.)
The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model to describe NFRs with  the following  concepts:  {\sc Business service}\footnotemark \footnotetext{\hl{We use the 
%{\sc 
small caps
%} 
font for referring to classes of a meta-model.}}, {\sc End consumer}, {\sc Requirement}, {\sc Use Case}, {\sc Composite Use Case}, {\sc Non-\-Func\-tion\-al Requirement}, {\sc Non-Functional} {\sc At\-tri\-bute} and {\sc Constraint}.
An {\sc End Consumer} is represented by an {\sc Actor}.
An  {\sc Actor} is related to {\sc Use Cases} (from the original UML definition), as a {\sc  Composite Use Case} is a set of actions performed by the system which can be broken into different {\sc Use Cases}.
{\sc  Business Service} aggregates several {\sc Use Cases},  a service can be expressed by one or more use cases.
 The {\sc Business Collaborator} concept is represented  by {\sc Packages}.
A {\sc Business Collaborator} denotes an external service or a system that interacts with the application that is being modeled.
Each {\sc Business Collaborator} combines the features described in each  {\sc Package}.
%Thus, the service functions can be specified as being grouped into packages.

@@

 The {\sc Non-functional requirement} and {\sc Non-functional Attribute} concepts are represented by {\sc Use Cases} and {\sc Constraints}.
A {\sc Use Case} may have several {\sc Constraints}. Each {\sc Constraint} has a name, description, and a flag to indicate that it has to be  dynamically checked.
Each {\sc Con\-straint} is represented as a stereotyped ({\sf constraint})\footnotemark \footnotetext{\hl{We use the 
%{\sf 
sans serif
%}
 font for referring to classes defined using a meta-model.}} use case.
There are three kinds of constraints:
\textit{(i)} A data pair ({\sc Value Constraint}), represented as the stereotype ``value'';
\textit{(ii)} Business rules (\textsc{Business Constraint}), represented as the stereotype {\sf ``business''}; and
\textit{(iii)} An {\sc Exceptional Behavior} constraint, represented by the stereotype {\sf ``exceptional\_behavior''}.
The use of these constraints is shown in the next example.

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic2}
Figure~\ref{fig:CIM:piusecasetpm} shows a $\pi$-UseCase model for our example.
We consider that, besides the service composition for implementing the application, it is necessary to model  other requirements that represent the \textit{(i)} conditions imposed by service usage -- for example, the fact that both Facebook and Twitter require authentication in order to call their methods for updating the wall; \textit{(ii)}  conditions stemming from the business rules of the application logic, (e.g., the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated).

@@

In Figure \ref{fig:e-scomposition-metamodel}, the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model
provides meta-classes to represent workflows%\footnotemark \footnotetext{\hl{Workflows will be transformed into (implemented) service compositions.}} 
 that model  business processes.
The \textit{$\pi$-Serv\-ice\-Com\-po\-si\-tion} meta-model extends the UML activity  meta-model with the concept of  \textit{A-Policy}
to group contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a single policy.
 This meta-model defines:

@@

\item In order to represent constraint types associated to \hl{service compositions}, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
%\martin{Placido: Mudar as cores... } 
 meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules.
The conditions of each rule are evaluated.
In case of failure, the actions of the rule will be performed.
The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.

@@

\begin{example}[To Publish Music \textit{(cont)}]\label{ex:toPublicMusic4}
To illustrate the use of the $\pi$-Service Composition meta-model, we define a model for the ``To Publish Music'' scenario (Figure \ref{fig:servicecompositionmodel}).
 We model a business process with three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
The {\em Publish Music} activity calls the {\em Facebook} and {\em Twitter} services.
Both {\em Facebook} and {\em Twitter} require authentication.
Two authentication policies are required, one for {\em Twitter} and another for {\em Facebook}.
%{\color{red} To be completed!.}
In this model, there are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook}).
% \footnotemark \footnotetext{\hl{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}}).
The model also shows the business process of the application that consists of three service activities: {\em Listen Music}, {\em Publish Music} and {\em Confirmation}.
Note that  the activity {\em Publish Music} calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.
Both {\em Facebook} and {\em Twitter} services require authentication in order to execute methods that read and update the user data.
%A call to such services must be part of the authentication protocol required by these services.
In the example, we  associate two authentication policies, one for the open authentication protocol, represented by the class {\sf\small OAuthPolicy} at {\em Twitter}, associated to the activity  {\sf\small UpdateTwitter} (see Figure \ref{fig:servicecompositionmodel}).
In the same way, the {\em Facebook} class {\sf\small HTTPAuthPolicy}, for the http authentication protocol is associated to the activity {\sf\small UpdateFacebook}.
The {\em A-policy} {\sf\small OAuthPolicy} has a variable {\sf\small Token}, used to store the authentication token provided by the service.
This variable is imported through the library {\sf\small OAuthPolicy.Token}.
The A-policy {\sf\small OAuthPolicy} defines two rules, both can be triggered by events of type {\sf\small ActivityPrepared}: ({\sf\small R$_1$}): If no token has been associated to the variable {\sf\small token}, then a token is obtained ; and ({\sf\small R$_2$}): if the token has expired, then it is renewed.
Notice that the code in the actions profits from the imported {\sf\small OAuthPolicy.Token} for transparently obtaining or renewing a token from a third party.
{\sf\small HTTPAuthPolicy} implements the HTTP-Auth protocol.
The A-policy imports an http protocol library and it has two variables {\sf\small username} and {\sf\small password}.
The event of type {\sf\small ActivityPrepared} is the triggering event of the rule {\sf\small R$_1$}.
On the notification of an event of that type, a credential is obtained using the username and password.
\end{example}

@@

\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03,ws-bpel2}.% or PEWS~\cite{Placido2010LTPD}.
%
PEWS~\cite{BHM06}%,Placido2010LTPD}
 is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:

@@

\subsection{Implementation}
We have developed tools for aiding the user to define and transform the models for all the levels, except for the CIM into PIM level, which should be manually performed.
Our tool is implemented as a series of Eclipse plug-ins: 
\begin{compactitem}
\item 	We  used the Eclipse Modeling Framework (EMF)\footnotemark \footnotetext{\hl{The EMF project is a modeling framework and code generation facility for building tools and other applications based on a structured data model.}}   to implement the  $\pi$-Service Composition and $\pi$-{\sc Pews}  meta-models. 
From these meta-models, we  developed plug-ins to support their graphical representation.

\item	 We used  ATL\footnotemark \footnotetext{\hl{http://eclipse.org/atl/. An ATL program is basically a set of rules that define how source model elements are matched and navigated to create and initialize the elements of the target models.}}
for implementing the  mappings between models.

\item 	We  used Acceleo\footnotemark \footnotetext{\hl{http://www.acceleo.org/pages/home/en}} to implement  the code generation plug-in to generate executable code. 
It takes a $\pi$-PEWS model and generates the code to be executed by the {\em
A-Policy} based service composition execution environment.
\end{compactitem}

@@

To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnotemark \footnotetext{\hl{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}}.%\martin{Is this a case study or a proof of concept?}
Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes, as well as to propose amendments to these practices.
The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.

@@

Figure~\ref{fig:BPMNmodel} shows the BPMN model
for the \FlyingPig\ scenario. This model is partitioned for better understanding the value exchanging process.
The model includes two pools representing the \textsl{ORCA} system and the \textsl{Brokers}.
Brokers have two lanes, the client \textsl{Company} and a \textsl{User}.
The user is a contact member of the company, who  coordinates the assessment process.
This process  involves other members of the company as well.
%
The risk assessment process starts after a request from a company.
This corresponds to the value model, in which the start stimulus triggers the whole process.
The request leads to the definition of a group of users that will answer questionnaires to evaluate risk.
Other tasks include amending a ``risky situation'' as well as producing evidence to show that a specific risk has been eliminated\footnotemark \footnotetext{\hl{Risky situations include  material facts such as not facilitating access to physically disabled people in a bank agency or having an unsecured access to the premises of the company.
They can be also inmaterial restrictions such as the protocol used for accessing data on the company's computer server.}}.

@@

The  PSM of our case study is given in Figure~\ref{fig:piPEWSFlyingPig}. This model is obtained by transforming the $\pi$-service composition model into a workflow.
Note that this workflow is implemented by using  \hl{BPEL constructors}~\cite{ws-bpel2} and \textit{policy} notions.

@@

Through the example we underlined that every application implements functional aspects that describe its application logic.
Recall that an application logic refers to routines that perform the activities to reach the application objective.
Also there are non-functional properties derived from NFR. 
They refer to strategies to be considered for the application execution such as security, isolation, adaptability, atomicity, and more.
These non-functional properties must be ensured at execution time.

\hl{In the context of service-oriented applications, ensuring non-functional properties is challenging owing to the nature of their components.
Those components are defined by APIs which do not necessarily export information about their internals.}
